Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-16 csf-2017/paper.tex
Almost at the page limit.

\begin{comment}
  \begin{figure}
    \centering
    \begin{mathpar}
      \Phi \rightarrow
      \begin{cases}
        \sem{\decode(n)}(\Phi) & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$}}\\
          \;\;\text{\footnotesize{and $\start \leq \addr \leq \addrend$ and $\memheap(\addr) = n$}}\\
          \;\;\text{\footnotesize{and $\perm \in \{ \exec,\rwx, \rwlx \}$} }
        \end{array}\\
        \failed                & \text{\footnotesize{otherwise}}
      \end{cases}
      \and \updatePcPerm{w} =
      \begin{cases}
        ((\exec,\gl),\start,\addrend,\addr) & \text{\footnotesize{if $w = ((\entry,\gl),\start,\addrend,\addr)$}}\\
        w & \text{\footnotesize{otherwise}}
      \end{cases}\\
      \and \stdUpdatePc{\Phi} =
      \begin{cases}
        \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$}}\\ 
          \;\;\text{\footnotesize{and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}}
        \end{array}\\
        \failed & \text{\footnotesize{otherwise}}
      \end{cases}
      \and \inferrule{ } { \sem{\fail}(\Phi) = \failed } \and
      \inferrule{ } { \sem{\halt}(\Phi) = (\halted,\memheap) } \and
      \inferrule{ } { \sem{\move{\refreg{r_1}}{r_2}}(\Phi) =
        \stdUpdatePc{}(\updateReg{r_1}{\memreg(r_2)}) } \and
      \inferrule{ \memreg(r_2) = \stdcap \\
        \perm \in \{ \rwx, \rwlx, \exec, \readwrite,
        \readwritel, \readonly \}\\
        \start \leq \addr \leq \addrend } {
        \sem{\load{\refreg{r_1}}{\refheap{r_2}}}(\Phi) =
        \updateReg{r_1}{\memheap(\addr)}} \and
      \inferrule{ \memreg(r_2) = \stdcap[\permp]\\
        \permp' = \decodePermPair{\memreg(r_2)}\\
        \permp'\sqsubseteq \permp } {
        \sem{\restricttwo{\refreg{r_1}}{r_2}}(\Phi) = \\
        \stdUpdatePc{}(\updateReg{r_1}{(\permp',\start,\addrend,\addr)})
      } \and \inferrule{ \memreg(r_2) = ((\_,\_),\_,\_,\addr) } {
        \sem{\geta{\refreg{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateReg{r_1}{\addr}}} \and \inferrule{
        \var{newPc} = \updatePcPerm{\memreg(\lv)} } {
        \sem{\jmp{\lv}}(\Phi) = \updateReg{\pcreg}{\var{newPc}} } \and
      \inferrule{ \memreg(r_1) = \stdcap\\
        \start \leq \addr \leq \addrend\\
        \perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}\\
        \var{w} = \memreg(r_2) \\
        \var{w} = ((\_,\local),\_,\_,\_)\Rightarrow\perm \in
        \{\rwlx,\readwritel \} } {
        \sem{\store{\refheap{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateHeap{\addr}{\var{w}}} } \and
    \end{mathpar}
    \caption{An excerpt from the operational semantics.}
    \label{fig:op-sem}
  \end{figure}
\end{comment}
\figurename~\ref{fig:op-sem} shows an excerpt of the operational semantics for a
number of representative instructions. Essentially, a configuration $\Phi$ will
either decode and execute the instruction at $\memreg(\pcreg)$ if it is
executable and the current address is in the valid range or fail otherwise. The
table in the figure shows for a given instruction $i$ the result of executing it
in configuration $\Phi$. The $\fail$ and $\halt$ instructions obviously fail and
halt respectively. The $\move{}{}$ instruction simply modifies the register file
as requested and updates the $\pcreg$ to the next instruction using the
meta-function $\stdUpdatePc{}$.

The $\load{}{}$ instruction loads the contents of the requested memory location
into the appropriate register, but only if the capability has appropriate
authority (i.e.\ read permission and an appropriate range). The $\restricttwo{}{}$
instruction will update a capability's permissions and global/local tag in
the register file, but requires that the new permissions are weaker than the
original and does not turn local capabilities into global ones. The $\geta{}{}$
instruction queries the current address of a capability and stores it in another
register.

The $\jmp{}$ instruction updates the program counter to a requested location,
but it is complicated by the presence of \emph{enter capabilities}, modeled
after the M-Machine's~\cite{Carter:1994:HSF:195473.195579}. Enter capabilities
cannot be used to read, write or execute and their current address and range
cannot be modified. They can only be used to jump to, but when that happens,
their permission is updated to $\exec$. This special behavior makes them
suitable for representing a kind of closures: an opaque package containing a
piece of code together with local encapsulated state. Such a package can be
built as an enter capability $c = ((\entry,\gl),\start,\addrend,\addr)$ where
the range $[\start,\addr-1]$ contains local state (data or capabilities) and
$[\addr,\addrend]$ contains instructions. The package is opaque to an adversary
holding $c$ but when $c$ is jumped to, the instructions can start executing and
have access to the local data through the updated version of $c$ that is then in
$\pcreg$\footnote{Note that the capability in the $\pcreg$-register can be used
  to load from.}.
% \lau{As we
%   talked about what is described here, is not how we make closures in the TR, so
%   we should consider whether we should refrain from using that word here. }

>>>>

\begin{comment}
  \begin{figure}
    \centering
    \begin{mathpar}
      \Phi \rightarrow
      \begin{cases}
        \sem{\decode(n)}(\Phi) & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$}}\\
          \;\;\text{\footnotesize{and $\start \leq \addr \leq \addrend$ and $\memheap(\addr) = n$}}\\
          \;\;\text{\footnotesize{and $\perm \in \{ \exec,\rwx, \rwlx \}$} }
        \end{array}\\
        \failed                & \text{\footnotesize{otherwise}}
      \end{cases}
      \and \updatePcPerm{w} =
      \begin{cases}
        ((\exec,\gl),\start,\addrend,\addr) & \text{\footnotesize{if $w = ((\entry,\gl),\start,\addrend,\addr)$}}\\
        w & \text{\footnotesize{otherwise}}
      \end{cases}\\
      \and \stdUpdatePc{\Phi} =
      \begin{cases}
        \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$}}\\ 
          \;\;\text{\footnotesize{and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}}
        \end{array}\\
        \failed & \text{\footnotesize{otherwise}}
      \end{cases}
      \and \inferrule{ } { \sem{\fail}(\Phi) = \failed } \and
      \inferrule{ } { \sem{\halt}(\Phi) = (\halted,\memheap) } \and
      \inferrule{ } { \sem{\move{\refreg{r_1}}{r_2}}(\Phi) =
        \stdUpdatePc{}(\updateReg{r_1}{\memreg(r_2)}) } \and
      \inferrule{ \memreg(r_2) = \stdcap \\
        \perm \in \{ \rwx, \rwlx, \exec, \readwrite,
        \readwritel, \readonly \}\\
        \start \leq \addr \leq \addrend } {
        \sem{\load{\refreg{r_1}}{\refheap{r_2}}}(\Phi) =
        \updateReg{r_1}{\memheap(\addr)}} \and
      \inferrule{ \memreg(r_2) = \stdcap[\permp]\\
        \permp' = \decodePermPair{\memreg(r_2)}\\
        \permp'\sqsubseteq \permp } {
        \sem{\restricttwo{\refreg{r_1}}{r_2}}(\Phi) = \\
        \stdUpdatePc{}(\updateReg{r_1}{(\permp',\start,\addrend,\addr)})
      } \and \inferrule{ \memreg(r_2) = ((\_,\_),\_,\_,\addr) } {
        \sem{\geta{\refreg{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateReg{r_1}{\addr}}} \and \inferrule{
        \var{newPc} = \updatePcPerm{\memreg(\lv)} } {
        \sem{\jmp{\lv}}(\Phi) = \updateReg{\pcreg}{\var{newPc}} } \and
      \inferrule{ \memreg(r_1) = \stdcap\\
        \start \leq \addr \leq \addrend\\
        \perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}\\
        \var{w} = \memreg(r_2) \\
        \var{w} = ((\_,\local),\_,\_,\_)\Rightarrow\perm \in
        \{\rwlx,\readwritel \} } {
        \sem{\store{\refheap{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateHeap{\addr}{\var{w}}} } \and
    \end{mathpar}
    \caption{An excerpt from the operational semantics.}
    \label{fig:op-sem}
  \end{figure}
\end{comment}
\figurename~\ref{fig:op-sem} shows an excerpt of the operational semantics for a
few representative instructions. Essentially, a configuration $\Phi$ either
decodes and executes the instruction at $\memreg(\pcreg)$ if it is executable
and its address is in the valid range or otherwise fails. The table in the
figure shows for instructions $i$ the result of executing them in configuration
$\Phi$. $\fail$ and $\halt$ obviously fail and halt respectively. $\move{}{}$
simply modifies the register file as requested and updates the $\pcreg$ to the
next instruction using the meta-function $\stdUpdatePc{}$.

The $\load{}{}$ instruction loads the contents of the requested memory location
into a register, but only if the capability has appropriate authority (i.e.\
read permission and an appropriate range). $\restricttwo{}{}$ updates a
capability's permissions and global/local tag in the register file, but only if
the new permissions are weaker than the original. It also never turns local
capabilities into global ones. $\geta{}{}$ queries the current address of a
capability and stores it in a register.

The $\jmp{}$ instruction updates the program counter to a requested location,
but it is complicated by the presence of \emph{enter capabilities}, modeled
after the M-Machine's~\cite{Carter:1994:HSF:195473.195579}. Enter capabilities
cannot be used to read, write or execute and their address and range cannot be
modified. They can only be used to jump to, but when that happens, their
permission changes to $\exec$. They can be used to represent a kind of closures:
an opaque package containing a piece of code together with local encapsulated
state. Such a package can be built as an enter capability $c =
((\entry,\gl),\start,\addrend,\addr)$ where the range $[\start,\addr-1]$
contains local state (data or capabilities) and $[\addr,\addrend]$ contains
instructions. The package is opaque to an adversary holding $c$ but when $c$ is
jumped to, the instructions can start executing and have access to the local
data through the updated version of $c$ that is then in $\pcreg$\footnote{Note
  that the capability in the $\pcreg$-register can be used to load from.}.
% \lau{As we
%   talked about what is described here, is not how we make closures in the TR, so
%   we should consider whether we should refrain from using that word here. }
